noam_data_exploration_v1

### 2024-07-17 noam_data_exploration_v3.Rmd
Added Anderson et al. 2013 portfolio effects at transect level

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(vegan)
library(glmmTMB)
library(car)
library(effects)
library(DHARMa)
library(stringr)
library(MuMIn)
library(codyn)
library(nlme)
library(mgcv)
library(emmeans)
```

@@

meta_cols <- c(colnames(data[1:7] ), "Island_side", "Island_side_habitat")
species_cols <- colnames(data[8:77])

@@

```{r turnover}
# the turnover function calculates turnover as the proportion or species that either appear or disappear between timepoints
total_turnover<-
  turnover(
  df = long_data_sub,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Percent_cover",
  replicate.var = "Location",
  metric = "total"
)

@@

@@

alpha_diversity$Cover <- rowSums(wide_transect_level[5:74])/100

diversity_stability <- alpha_diversity %>%
  dplyr::group_by(Site, Habitat, Site_habitat, Transect) %>%
  dplyr::summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Richness_cv = CV(Richness),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()),
            Shannon_cv = CV(Shannon),
            Cover_mean = mean(Cover),
            Cover_se = sd(Cover)/sqrt(n()),
            Cover_cv = CV(Cover))

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(vegan)
library(glmmTMB)
library(car)
library(effects)
library(DHARMa)
library(stringr)
library(MuMIn)
library(codyn)
library(nlme)
library(mgcv)
library(emmeans)
library(ecofolio) # possible other method for examining portfolio effects 
```

@@

meta_cols <- c(colnames(data[1:7] ), "Island_side", "Island_side_habitat", "Stratum")
species_cols <- colnames(data[8:77])

@@

```{r turnover, eval = FALSE}
# the turnover function calculates turnover as the proportion or species that either appear or disappear between timepoints
total_turnover<-
  turnover(
  df = long_data_sub,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Percent_cover",
  replicate.var = "Location",
  metric = "total"
)

@@

```{r transect level portfolio effect}
# using Anderson et al. 2013 definitions of portfolio effects - relevant fxns in ecofolio packages
wide_transect_level$Stratum <- as.factor(paste(wide_transect_level$Site, "-", wide_transect_level$Habitat, "-", wide_transect_level$Transect))
x <- subset(wide_transect_level, Stratum == "LTER 1 - Backreef - 1")

y <- x[ , -which(names(x) %in% meta_cols)]
y <- y[, which(colSums(y) != 0)]
z <- pe_avg_cv(y, ci = TRUE, detrending = "loess_detrended")

transect_portfolio_columns <- c("Stratum", "Portfolio_effect") # names of columns for dataframe
transect_level_portfolio <- 
  data.frame(matrix(nrow = length(unique(wide_transect_level$Stratum)), 
                    ncol = length(transect_portfolio_columns))
             )
colnames(transect_level_portfolio) <- transect_portfolio_columns # change column 
transect_level_portfolio$Stratum <- levels(wide_transect_level$Stratum) # populate

for(i in 1:length(unique(unique(wide_transect_level$Stratum))) ){ # for each  transect
  # create a new dataframe for current transect
  newData <- subset(wide_transect_level, Stratum == levels(wide_transect_level$Stratum)[i])
  current_matrix <- newData[ , -which(names(newData) %in% meta_cols)] # remove metadata columns
  if(sum(current_matrix) == 0){ # if this was a 0 algae transect
    transect_level_portfolio[i, 2] <- NA # set portfolio effects to 0
  }else{ # otherwise
    current_matrix <- current_matrix[, which(colSums(current_matrix) != 0)] # remove irrelevant algae

    transect_level_portfolio[i, 2] <- pe_avg_cv(current_matrix) # calculate portfolio effects for current time series
  }

alpha_diversity$Cover <- rowSums(wide_transect_level[ , -which(names(wide_transect_level) %in% meta_cols)])/100
alpha_diversity$Stratum <- paste(alpha_diversity$Site, "-", alpha_diversity$Habitat, "-", alpha_diversity$Transect)
lm_synchrony_transect <- codyn::synchrony(df = transect_level,
                                   time.var = "Year",
                                   species.var = "Taxa",
                                   abundance.var = "Mean_cover",
                                   metric = "Loreau",
                                   replicate.var = "Stratum") # Loreau and Mazancourt synchrony
colnames(lm_synchrony_transect) = c("Stratum", "Loreau_synchrony") # rename cols for merge later

gross_synchrony_transect <- codyn::synchrony(df = transect_level,
                                   time.var = "Year",
                                   species.var = "Taxa",
                                   abundance.var = "Mean_cover",
                                   metric = "Gross",
                                   replicate.var = "Stratum") # lots of warnings in this one - examine further later
colnames(gross_synchrony_transect) = c("Stratum", "Gross_synchrony") # rename cols for merge later

# stability_transect <- community_stability(df = transect_level,
#                                           time.var = "Year",
#                                           abundance.var = "Mean_cover",
#                                           replicate.var = "Stratum") # using codyn measure of stability
# NOTE: codyn community_stability() is just the inverse of CV
diversity_stability <- alpha_diversity %>%
  dplyr::group_by(Site, Habitat, Site_habitat, Transect, Stratum) %>%
  dplyr::summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Richness_cv = CV(Richness),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()),
            Shannon_cv = CV(Shannon),
            Shannon_stability = 1/CV(Shannon), # Stability as defined by codyn and many others (including Pete in 2022 paper)
            Cover_mean = mean(Cover),
            Cover_se = sd(Cover)/sqrt(n()),
            Cover_cv = CV(Cover),
            Cover_stability = 1/CV(Cover)) # as above

diversity_stability <- merge(diversity_stability, lm_synchrony_transect, by = "Stratum") # add in loreau and de Mazancourt synchrony
diversity_stability <- merge(diversity_stability, gross_synchrony_transect, by = "Stratum")
diversity_stability <- merge(diversity_stability, transect_level_portfolio, by = "Stratum") # add in portfolio effects
pairs(~ Cover_mean + Cover_cv + Cover_stability + 
        Shannon_mean + Shannon_cv + Shannon_stability + 
        Loreau_synchrony + Gross_synchrony + Portfolio_effect, data = diversity_stability)

```{r diversity stability quadrat level}

@@
